---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# QC plots of mapping metrics


This notebook illustrates the use of the ggplot2 package for producing QC plots mapping metrics


### Load packages

```{r}
library(tidyverse)
```

### Import Counts

```{r}
procdir <- normalizePath("./proc/")
qcfile <- file.path(procdir, "PRJNA668393-qcdata.RDS")

PRJNA668393qcData <- readRDS(qcfile)

tools::md5sum(qcfile)
```

Note that the libraries are across the rows (the variables are across the columns)

```{r}
PRJNA668393qcData
```

## Visualize proportion of mapped reads


### Simple plot

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped)) + 
    ggplot2::geom_point()
```

### Add a theme

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped)) + 
    ggplot2::geom_point() +
    ggplot2::theme_bw() +
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Define common theme


One can define a common theme to be applied to multiple figures

```{r}
mytheme <- theme_bw() + ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Annotate treatment information with colors

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped, col = treatment)) + 
    ggplot2::geom_point() + 
    mytheme
```

### Annotate treatment information with colors and cell line levels with shape

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped, col = treatment, shape = cell_line)) + 
    ggplot2::geom_point() + 
    mytheme
```

### Annotate treatment information with colors and cell line levels with shape(with additional customization)

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped, col = treatment, shape = cell_line)) +
    ggplot2::geom_point(size = 3) +
    ggplot2::labs(col = "Treatment") +
    ggplot2::labs(shape = "Cell Line") +
    ggplot2::xlab("Run Label") + 
    ggplot2::ylab("Total reads (depth)") +
    ggplot2::scale_colour_manual(values = c("red3", "blue3", "green3")) +
    mytheme
```

This is a variation of the previous figure where the legends have been combined

```{r}
PRJNA668393qcData %>%
    dplyr::mutate(grouplab = factor(paste(cell_line, treatment, sep = ", "))) %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped, col = grouplab, shape = grouplab)) +
    ggplot2::geom_point(size = 3) +
    ggplot2::scale_colour_manual(name = "Cell Line, Treatment", values=c("blue3","red3","blue3","red3")) +   
    ggplot2::scale_shape_manual(name = "Cell Line, Treatment", values=c(19, 19, 17, 17)) +
    ggplot2::xlab("Run Label") + 
    ggplot2::ylab("Proportion of Uniquely Mapped Reads") +
    mytheme
    
```

## Visualize depth

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = depth)) +
    ggplot2::geom_point(size = 3) +
    mytheme
```

A slightly customized version of the previous plot suggesting that the UV2 cell lines were sequenced at higher depth

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = depth, col = treatment, shape = cell_line)) +
    ggplot2::geom_point(size = 3) +
    ggplot2::labs(col = "Treatment") +
    ggplot2::labs(shape = "Cell Line") +
    ggplot2::xlab("Run Label") + 
    ggplot2::ylab("Total reads (depth)") +
    ggplot2::scale_colour_manual(values = c("red3", "blue3", "green3")) +
    mytheme
```

## Visualize proportion of uniquely mapped reads with facets

```{r}
PRJNA668393qcData %>% ggplot(aes(x=Run, y=prop.mapped, shape =cell_line, color=treatment))+
    ggplot2::geom_point(size = 3) +
    ggplot2::labs(col = "Treatment") +
    ggplot2::labs(shape = "Cell Line") +
    ggplot2::scale_colour_manual(values = c("red3", "blue3", "green3")) +
    ggplot2::facet_grid(treatment~ cell_line, drop=TRUE, scales="free_x", space="free") +
    mytheme
```

```{r}
sessionInfo()
```
