---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Query genomic annotation using the biomaRt package from Bioconductor

```{r}
library(biomaRt)
library(tidyverse)
```

```{r}
mymart <- useMart("ENSEMBL_MART_ENSEMBL", dataset="mmusculus_gene_ensembl", version = "Ensembl Genes 104")
```

## Example: query entrez id, and MGI id and symbol for ensembl ids


### List filters

```{r}
listFilters(mymart) %>% dplyr::filter(stringr::str_detect(name, "ensembl"))
```

### List attributes

```{r}
listAttributes(mymart) %>% dplyr::filter(str_detect(name, "ensembl|entrez|mgi"))
```

### sample query for two genes


Query by ensembl id

```{r}
biomaRt::getBM(
    filters = "ensembl_gene_id",
    attributes=c("ensembl_gene_id", "entrezgene_id","mgi_id", "mgi_symbol"),
    values=c("ENSMUSG00000023951", "ENSMUSG00000041872"),
    mart=mymart
)
```
Conversely, query using entrez ids

```{r}
biomaRt::getBM(
    filters = "entrezgene_id",
    attributes=c("ensembl_gene_id", "entrezgene_id","mgi_id", "mgi_symbol"),
    values=c(22339, 257630),
    mart=mymart
)
```





## Query gene ids for pathway analysis


A pathway analysis is by deisgn limited to the genes annotated in pathways. Accordingly, can limit our attention to this subset of genes.


### Import Pathways

```{r}
annodir <- "./resource/"
annofile <- file.path(annodir, "Mm.h.all.v7.1.entrez.rds")

tools::md5sum(annofile)

Mm.H <- readRDS(annofile)
```

### Query annotation for genes annotated in the pathways

```{r}
biomaRt::getBM(
    filters = "entrezgene_id",
    attributes=c("chromosome_name", "start_position", "end_position" , "strand" , "ensembl_gene_id", "entrezgene_id","mgi_id", "mgi_symbol"),
    values=as.integer(unique(unlist(Mm.H))),
    mart=mymart
) -> pathwayids
```

```{r}
pathwayids %>% head
```

### 


Number of pathways ids with no query

```{r}
length(setdiff(as.integer(unlist(Mm.H)), pathwayids[["entrezgene_id"]]))
```

Pathway ids with non-unique ensembl id

```{r}
pathwayids %>% dplyr::group_by(ensembl_gene_id) %>% dplyr::filter(n()>1) %>% arrange(ensembl_gene_id)
```

```{r}
## 
biomaRt::getBM(
    filters = "entrezgene_id",
    attributes=c("ensembl_gene_id", "entrezgene_id","mgi_id", "mgi_symbol"),
    values=c(19047,434233),
    mart=mymart
)          
```

```{r}
## 
biomaRt::getBM(
    filters = "entrezgene_id",
    attributes=c("ensembl_gene_id", "entrezgene_id","mgi_id", "mgi_symbol"),
    values=c(100041273, 100042503, 68194),
    mart=mymart
)          
```

```{r}
56066

biomaRt::getBM(
    filters = "entrezgene_id",
    attributes=c("ensembl_gene_id", "entrezgene_id","mgi_id", "mgi_symbol"),
    values=c(56066),
    mart=mymart
) 
```

## Import ensembl ids from study data

```{r}
listFilters(mymart)[grep("entrez",listFilters(mymart)[,1]),]
```

## Query annotation data from MGI (http://www.informatics.jax.org)

```{r}
paste0(rep("c",15), collapse = "")
```

```{r}
mgifile <- "http://www.informatics.jax.org/downloads/reports/MGI_Gene_Model_Coord.rpt"

readr::read_tsv(mgifile) -> mgidat
```

```{r}
mgidat 
```

## Get rid of genes listed as "null"

```{r}
mgidat %>% 
    dplyr::filter(`6. Entrez gene id` != "null" & `11. Ensembl gene id` != "null") ->
    mgidat

mgidat
```

### Check  if ids are unique`

```{r}
mgidat %>% 
    dplyr::group_by(`6. Entrez gene id`) %>%
    dplyr::filter(n()>1)
```

```{r}
mgidat %>% 
    dplyr::group_by(`11. Ensembl gene id`) %>%
    dplyr::filter(n()>1)
```

### Number of pathway genes  not in annotation file

```{r}
length(setdiff(unlist(Mm.H), mgidat[["6. Entrez gene id"]]))
```

### Set location of analysis results

```{r}
procdir <- "./proc"

addfile <- file.path(procdir, "PRJNA668393-DESeq2-additive.RDS")
multfile <- file.path(procdir, "PRJNA668393-DESeq2-multiplicative.RDS")

tools::md5sum(addfile)
tools::md5sum(multfile)
```

### 

```{r}
### The ensembl ids from the gtf file end with version numbers
### One can use str_remove to remove the ensemble version from id
stringr::str_remove("ENSMUSG12345.01",".[0-9][0-9]{0,2}$")
stringr::str_remove("ENSMUSG12345.1",".[0-9][0-9]{0,2}$")


```

### Create a named vector (using entrez ids) for statistics from additive model

```{r}
### drop genes with na statistics
### clean ensembl id by dropping version number
### Merge the results by the clean ensemble id using an inner join
### Only keep 

readRDS(addfile) %>%
    DESeq2::results(tidy = TRUE) %>%
    tidyr:: drop_na(stat) %>%
    dplyr::mutate(ensid = str_remove(row, ".[0-9][0-9]{0,2}$")) %>%
    dplyr::inner_join(mgidat, by = c("ensid" = "11. Ensembl gene id")) %>%
    dplyr::filter(`6. Entrez gene id` %in% unique(unlist(Mm.H))) %>% 
    dplyr::pull(stat, name = "6. Entrez gene id") ->
    addstat
```

#### spot check the conversion for three genes

```{r}
idx <- sample(1:length(addstat), 3)
idx
```

```{r}
mgidat %>% 
    dplyr::filter(`6. Entrez gene id` %in% names(addstat[idx])) %>%
    dplyr::select(`6. Entrez gene id`, `11. Ensembl gene id`) ->
    testres
    
testres
```

```{r}
addstat[idx]
readRDS(addfile) %>%
    DESeq2::results(tidy = TRUE) %>%
    dplyr::filter(str_detect(row, paste0(testres[["11. Ensembl gene id"]], collapse = "|")))
```

### Create a named vector (using entrez ids) for statistics from additive model

```{r}
### drop genes with na statistics
### clean ensembl id by dropping version number
### Merge the results by the clean ensemble id using an inner join
### Only keep 

readRDS(multfile) %>%
    DESeq2::results(tidy = TRUE) %>%
    tidyr:: drop_na(stat) %>%
    dplyr::mutate(ensid = str_remove(row, ".[0-9][0-9]{0,2}$")) %>%
    dplyr::inner_join(mgidat, by = c("ensid" = "11. Ensembl gene id")) %>%
    dplyr::filter(`6. Entrez gene id` %in% unique(unlist(Mm.H))) %>% 
    dplyr::pull(stat, name = "6. Entrez gene id") ->
    multstat
```

#### Spot check

```{r}
idx <- sample(1:length(addstat), 3)
idx
```

```{r}
mgidat %>% 
    dplyr::filter(`6. Entrez gene id` %in% names(multstat[idx])) %>%
    dplyr::select(`6. Entrez gene id`, `11. Ensembl gene id`) ->
    testres
    
testres
```

```{r}
multstat[idx]
readRDS(addfile) %>%
    DESeq2::results(tidy = TRUE) %>%
    dplyr::filter(str_detect(row, paste0(testres[["11. Ensembl gene id"]], collapse = "|")))
```

```{r}
outdir <- "./proc"

addstatfile <- file.path(outdir, "PRJNA668393-DESeq2-addstat.RDS")
multstatfile <- file.path(outdir, "PRJNA668393-DESeq2-multstat.RDS")

saveRDS(addstat, addstatfile)
saveRDS(multstat, multstatfile)

tools::md5sum(addstatfile)
tools::md5sum(multstatfile)
```

```{r}
sessionInfo()
```

```{r}

```
