# R libraries and Bioconductor

## Packages and Libraries

R is at heart a collection of 'packages'. There is a 'base' system that contains the truly basic commands, such as the assignment operator `->` or the command to create a vector. In addition to that, there are 'standard R' packages that are included when you install the R kernel (in the Jupyter notebook), or 'R' as a program to run either at the command line or with Rstudio. (I've shown some examples of these different ways to run R in class).

### Libraries

Many packages, even those included in [standard R] (https://www.r-project.org/), will need to be 'loaded' to be used. In other words, they exist on your computer (or in your container), but the R kernel doesn't know about them. This is because if it did, R would be using computer memory (RAM) to remember all their functions and variables. If all the available packages were loaded, you might not have any RAM left!

A consequence of this is that you often have to tell R explicitly that you want to use a particular package. You do that using `library`. We'll use the `mtcars` data set for some illustration.




```{r}
head(mtcars)
```

There is a cool R function that will allow us to look at some random rows from a data frame. It's called `sample_n`. Let's try it:


```{r}
sample_n(titanic, 10)
```

Oops. It turns out `sample_n` is in the dplyr package. It's installed in your container - but R doesn't know that! Let's tell R we want to use it:


```{r}
library(dplyr)
```


```{r}
sample_n(mtcars, 10)
```

### Installed and installing packages

Now, `dplyr` is actually not part of standard R. It's *installed* separately. There are a multitude of R packages out there. Anyone can write one (yes, even you!!!). They are shared with the public using the [CRAN archive.] (https://cran.r-project.org/) In order to be listed in CRAN, packages need to meet specific criteria for documentation purposes, testing, etc.

You can check to see what packages are installed using `installed.packages()`


```{r}
installed.packages()
```

You can install new packages using the command `install.packages()`



```{r}
system("mkdir -p ~/R")
install.packages("auk", lib = "~/R")
```


```{r}
remove.packages("auk", lib = "~/R")

```

## BioConductor

CRAN is home to many, many R packages. But there is a whole other world out there when it comes to bioinformatics in R. It's called [BioConductor](https://bioconductor.org/). BioConductor is a comprehensive toolkit for all things having to do with high-throughput sequencing data processing and analysis. In this course, we will use the BioConductor package `DESeq2` to perform differential expression analysis. It's the end of the pipeline, after QC, clipping and trimming, aligning and counting. 

### Installing BioConductor packages

BioConductor has it's own installation procedure (and it's own criteria for documentation, testing, etc.) - separate from CRAN. Let's have a look at the page for [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html)


```{r}
BiocManager::install("DESeq2", lib = "~/R")
```

### DESeq2 and S4 Objects

We'll walk through an example using a sample data set called 'airway'. Airway is an object of type 'SummarizedExperiment'. This kind of object is the basis for many objects used in Bioconductor packages.


```{r}
library("airway")
data("airway")
se <- airway
```


```{r}
str(se)
```

[This tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html) gives a great introduction to the SummarizedExperiment object. We'll take a peek, and then move on to DESeq2 


```{r}
assays(se)
```


```{r}
assays(se)$counts
```


```{r}
rowRanges(se)
```


```{r}
# Sample info

colData(se)
```


```{r}
metadata(se)
```


```{r}
# Just a list - we can add elements

metadata(se)$formula <- counts ~ dex + albut

metadata(se)
```


```{r}
# subset the first five transcripts and first three samples
se[1:5, 1:3]
```


```{r}
assays(se[1:5,1:3])$counts

```


```{r}
library("DESeq2")


dds <- DESeqDataSet(se, design = ~ cell + dex)
dds


```


```{r}
# remove rows with less than 10 total transcripts

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```


```{r}
ddsSEQ <- DESeq(dds)
res <- results(ddsSEQ)
res
```

```{r}

```
