---
title: QC plots
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

### Load packages

```{r}
library(tidyverse)
```

### Import Counts

```{r}
procdir <- normalizePath("./proc/")
qcfile <- file.path(procdir, "PRJNA668393-qcdata.RDS")

PRJNA668393qcData <- readRDS(qcfile)

tools::md5sum(qcfile)
```

### Specify a theme for the plots

```{r}
mytheme <-  ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Visualize proportion of mapped reads


### Simple plot

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped)) + 
        ggplot2::geom_point()
```

### Add a theme

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped)) + 
        ggplot2::geom_point() +
            ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Annotate treatment information with colors

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped, col = treatment)) + 
        ggplot2::geom_point() + 
           ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Annotate treatment information with colors and cell line levels with shape

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped, col = treatment, shape = cell_line)) + 
        ggplot2::geom_point() + 
            ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Annotate treatment information with colors and cell line levels with shape(with additional customization)

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = prop.mapped, col = treatment, shape = cell_line)) +
        ggplot2::geom_point(size = 3) +
            ggplot2::labs(col = "Treatment") +
                ggplot2::labs(shape = "Cell Line") +
                    ggplot2::xlab("Run Label") + 
                        ggplot2::ylab("Proportion of Uniquely Mapped Reads") +
                            ggplot2::scale_colour_manual(values = c("red3", "blue3", "green3")) +
                                ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Visualize depth

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = Run, y = depth, col = treatment, shape = cell_line)) +
        ggplot2::geom_point(size = 3) +
            ggplot2::labs(col = "Treatment") +
                ggplot2::labs(shape = "Cell Line") +
                    ggplot2::xlab("Run Label") + 
                        ggplot2::ylab("Total reads (depth)") +
                            ggplot2::scale_colour_manual(values = c("red3", "blue3", "green3")) +
                                ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
PRJNA668393qcData %>%
    ggplot2::ggplot(aes(x = depth, y = prop.mapped, col = treatment, shape = cell_line)) +
        ggplot2::geom_point(size = 3) +
            ggplot2::labs(col = "Treatment") +
                ggplot2::labs(shape = "Cell Line") +
                    ggplot2::xlab("Run Label") + 
                        ggplot2::ylab("Proportion of Uniquely Mapped Reads") +
                                ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Visualize proportion of uniquely mapped reads with facets

```{r}
PRJNA668393qcData %>% ggplot(aes(x=Run, y=prop.mapped, shape =cell_line, color=treatment))+
    ggplot2::geom_point(size = 3) +
        ggplot2::labs(col = "Treatment") +
            ggplot2::labs(shape = "Cell Line") +
                ggplot2::scale_colour_manual(values = c("red3", "blue3", "green3")) +
                    ggplot2::facet_grid(treatment~ cell_line, drop=TRUE, scales="free_x", space="free") +
                        ggplot2::theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
sessionInfo()
```
